---
title: "Gov 50 Final Project"
author: "Lauren"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

## Project thoughts
Hello! I have changed my proposal and project trajectory. I wasn’t sure if I had to redo milestones 1 and 2, but decided to provide a brief summary. For my final project, I would like to explore the correlation between music and mental health, specifically whether listening to a certain type of music improves levels of depression.

## Proposal
Is there a correlation between frequency of a certain music genre listened to and depression levels? Do people who suffer with depression gravitate towards a specific type of music to improve mood and cognition? My hypothesis is that depression is positively correlated with listening to music with higher BPM and more uplifting/energizing lyrics.

Link to my data source: https://www.kaggle.com/datasets/catherinerasgaitis/mxmh-survey-results/data

## Data Visualization
```{r}
library(readr)
library(dplyr)
library(ggplot2)

# Read the data
mxmh_survey_results <- read_csv("~/Desktop/gov50/mxmh_survey_results.csv")
mxmh_survey_results

frequency_mapping <- c("Never" = 0, "Rarely" = 1, "Sometimes" = 2, "Often" = 3, "Very frequently" = 4)
mxmh_survey_results$`Frequency [Rock]` <- factor(mxmh_survey_results$`Frequency [Rock]`, levels = names(frequency_mapping))
mxmh_survey_results$`Frequency [Rock] Numeric` <- as.numeric(mxmh_survey_results$`Frequency [Rock]`)
mxmh_survey_results <- mxmh_survey_results |>
  filter(`Music effects` == "Improve")

plot <- ggplot(mxmh_survey_results, aes(x = Depression, y = `Frequency [Rock] Numeric`)) +
  geom_point(aes(color = as.factor(Depression)), alpha = 0.6) + # Use color to denote the different levels of depression
  geom_smooth(method = "lm", se = FALSE, color = "blue") + # Add a linear regression line without the confidence interval shading
  scale_y_continuous(limits = c(0, 4), breaks = 0:4, 
                     labels = c("Never", "Rarely", "Sometimes", "Often", "Very frequently")) + # Adjust the scale of y-axis
  labs(x = "Depression Level", y = "Frequency of Rock Music", 
       title = "Correlation between Depression and Frequency of Rock Music") +
  theme_minimal() 
plot
```

## Analysis
```{r}
fit.app <- lm(BPM ~ Depression, data = mxmh_survey_results)
fit.app
```
For a one-point increase in indicated depression level, BPM of music listened to increased by 0.59 points. This means that there is a slightly positive correlation between depression levels and BPM of music listened to.
